---
title: "Quant Methods_1"
output: pdf_document
---
Hit tab over an argument to pull up some helpful information

```{r}
dat<-read.csv("./inflammation-01.csv", header= F)
head(dat)
#summary--shows us what the first 6 ross of the data say-columns= differnt point in time and rows are patients
#To change the number of rows
head(dat, 10)
tail (dat)
class(dat)
#what is dat
```
R has some different classes-- dataframe() is one of the more common ones
4 most common classes of objects= data.farme(), vector(), matix(), list(). Lists are most flexible

Subsetting data-[row, column]
```{r}
dat[1,1]
dim(dat) 
# data dimensions- counting the rows and the columns
dat[30, 20]
```
Splicing- taking a big chunk of the data= Ex: rows 1-10/ finding some basic stats for the different patients
```{r}
dat[ 1:4, 1:10]
# we can also list what rows we are interested in- concatenate into a vector
c(3,2,4)
dat[c(3,8,37,56), c(10,14,29)]
#all columns from row 5
dat[5, ]
#all rows from column 16
dat[ ,16]
patient_1<- dat[1, ]
max(patient_1)
summary(dat[ , 7])
```
Apply function- basic way to apply any function to some data frame- apply to rows or columns
```{r}
avg_patient_inflammation <- apply(dat, 1, mean)
avg_day_inflammation <- apply(dat, 2, mean)
#std deviation of day
std_day_inflammation <- apply(dat, 2, sd)
#std deviation of patients
std_patient_inflammation <- apply(dat, 1, sd)
#point test
sd(dat[1,])
```
Plotting data
```{r}
plot(avg_day_inflammation)
?pch
# go back and look at PDF business
plot(1:40, avg_day_inflammation, xlab='day', ylab='inflammation', frame.plot=F, col='magenta', pch=19, cex= 2, axes=F, main='Practice Plot')
axis(side=1, cex.axos=3, lwd=4, padj=1.5)
axis(side=2, cex.axos=3, lwd=4, padj=1.5)
#commands for making types of files- need to put it somewhere- . = working directory
#output a bunch of  figures- run PDF 


```

